A new form of transcorrelated Hamiltonian inspired by range-separated DFT

### 2020-12-02 cusp_j/paper/draft.tex

\definecolor{darkgreen}{HTML}{009900}
\usepackage[normalem]{ulem}
\newcommand{\sphi}[1]{\hat{{\bf S}}_{#1}}
\newcommand{\overlap}[2]{\langle #1 | #2 \rangle}
\newcommand{\matelem}[3]{\langle #1 | #2 | #3 \rangle}
\newcommand{\deriv}[3]{\frac{\partial^{#3} #1}{\partial {#2}^{#3}}}
\newcommand{\bd}[1]{{\bf {#1}}}
\newcommand{\br}[0]{{\bf {r}}}
\newcommand{\bri}[1]{{\bf r}_{#1}}
\newcommand{\bs}[0]{{\bf {s}}}
\newcommand{\dr}[1]{\text{d}{\bf r_{#1}}}
\newcommand{\psiex}[0]{\Psi^{\text{ex}}_0}
\newcommand{\phiex}[0]{\Phi^{\text{ex}}_0}
\newcommand{\phimu}[0]{\Phi^{\text{ex},\mu}_0}
\newcommand{\phimub}[0]{\Phi^{\mathcal{B},\mu}_0}
\newcommand{\xhimub}[0]{X^{\mathcal{B},\mu}_0}
\newcommand{\psimub}[0]{\Psi^{\mathcal{B},\mu}_0}
\newcommand{\phiimub}[0]{\Phi^{\mathcal{B},\mu}_i}
\newcommand{\basis}[0]{\mathcal{B}}
\newcommand{\energyex}[0]{E^{\text{ex}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\identity}{\mathds{1}}
\newcommand{\mur}[1]{\mu({\bf r_{#1}})}
\newcommand{\muueg}{\mu_{\text{UEG}}}
\newcommand{\muuegav}{\langle \mu_{\text{UEG}}\rangle}
\newcommand{\mursc}{ \mu_{r_{s,c}}}
\newcommand{\murscav}{\langle \mu_{r_{s,c}}\rangle}
\newcommand{\mursclda}{\langle \mu_{r_{s,c}^{\text{UEG}}}\rangle}

@@

\maketitle
\section{Introduction}
One of the most challenging problem in modern computational chemistry is the accurate simulation of the electronic structure of atomic and molecular systems, which implies solving the Schroedinger equation for a general $N$-body system. 
At the heart of the complexity of such a task relies the rapidly prohibitive computational cost of the wave function methods (WFT), which is induced by the exponential growth with the system size of the Hilbert space involved in the linear eigenvalue problem to be solved.

In usual WFT, the computational bottleneck is mainly determined by two factors: the level of complexity of the wave function which is imposed by the amount of strong correlation effects in the system, and the convergence of the computed quantities with respect to the size of the one-electron basis used to project the Hamiltonian into a finite eigenvalue problem. 
Typically, strong correlation effects appear when the two-electron coulomb interaction can no longer be considered as a small perturbation over a mean field Hamiltonian. Treating such effects in a black box way involves the use of very flexible wave functions such as 
selected CI\cite{HurMalRan-JCP-73,bender,malrieu,buenker1,buenker-book,three_class_CIPSI,harrison,GinSceCaf-CJC-13, GinSceCaf-JCP-15, ShaHolJeaAlaUmr-JCTC-17}, 
%selected CI\cite{HurMalRan-JCP-73,bender,malrieu,buenker1,buenker-book}, 
full configuration Quantum Monte Carlo\cite{BooThoAla-JCP-09,BooAla-JCP-10,BooCleThoAla-JCP-11,GhaLozAla-JCP-19,VitAlaKat-JCTC-20} (FCI-QMC) or tensor product states\cite{ChaSha-AR-11,BaiRei-JCP-20}. These methods all have one thing in common: they manage to select, although through different mathematical forms, the dominant part of the exponentially growing Hilbert space that resulting from a $N$-electron problem among a given finite basis set.

Even if these selection schemes are extremely efficient to obtain a qualitative description of the wave function, a quantitative description is often out of reach as it implies the use of very large one-electron basis sets which are needed to recover the slowly converging correlation effects near the electron-electron coalescence point. 
These short-range correlation effects are mainly due to the diverging character of the coulomb potential, which induces a non analytical behaviour of the wave function at small inter-electronic distance: the famous electronic cusp originally derived by Kato\cite{Kat-CPAM-57}. As shown by the seminal works of Hylleraas \cite{Hyl-ZP-29} and further developed byKutzelnigg and coworkers, \cite{Kut-TCA-85,KutKlo-JCP-91, NogKut-JCP-94} the slow convergence of the correlation energy with respect to the size of the basis set is mainly due to the impossibility to represent the cusp in a finite basis set. 
To tackle this problem, the introduction of functions explicitly depending on the inter-electronic coordinate (F12) have emerged\cite{Ten-TCA-12,TenNog-WIREs-12,HatKloKohTew-CR-12, KonBisVal-CR-12, GruHirOhnTen-JCP-17, MaWer-WIREs-18}. 
In the F12 schemes, the geminal function depending on the inter-electronic coordinate is used to describe short-range correlation effects which are absent from the finite basis set. Thanks to strong orthogonality, 
all redundant correlation effects of the geminal function are projected out of the wave function, 
and therefore the geminal function only deals with correlation orthonormal to the basis set. 
The resulting F12 theories, mainly based on perturbation theory and coupled cluster theory, are therefore 
extremely close to their parent theory, with additional integrals two- and three-body terms resulting from the use of the geminal function. 
It is important to notice that because of the orthogonalization between the geminal and the basis set, 
the complexity of the $N$-body problem within a given basis set is essentially the same between a F12 theory and their usual parent. Therefore, no strong compression of the electronic wave function is obtained by the F12 theory, 
but a fast convergence toward the complete basis set limit (CBS) is nevertheless obtained.

An alternative point of view using a geminal function is the so-called transcorrelated (TC) approach where the full effect of the geminal function is incorporated into the calculation through a similarity transformation of the original Hamiltonian. 
The pioneers of such theory were Boys, Handy and co-workers\cite{BoyHan-PRSLA-69,BoyHanLin-1-PRSLA-69,BoyHanLin-2-PRSLA-69} where they derive the equations of the general form of the TC Hamiltonian which is non hermitian and contains explicitly three-body interaction. Despite the loss of variational principle due to the non hermitian nature of the TC Hamiltonian, the parameters optimized by Boys and Handy were mainly the orbitals of a single Slater determinant together with the mathematical form of the geminal function with which one performs the similarity transformation of the Hamiltonian to obtain the TC Hamiltonian.

Later on, Ten-No\cite{Ten-No-CPL-00} proposed to significantly change the paradigm of Boys and Handy: instead of using a single Slater determinant and optimizing the orbitals and the geminal function, one uses a more elaborate many body theory (Moller-Plesset at second order in Refs.\onlinecite{Ten-No-CPL-00,HinTanTen-JCP-01} and coupled cluster in later works\cite{HinTanTen-CPL-02}) with a fixed universal geminal function. Such a geminal function, developed as a linear combination of gaussian functions, is optimized in order to remove the divergent character of the coulomb potential within the TC framework. 
In Ref.\onlinecite{HinTanTen-JCP-01}, Ten-No \textit{et. al.} introduced a biorthogonal approach, 
which allows to treat the non-hermitian TC Hamiltonian in a many-body framework which is therefore more suited 
to develop many-body perturbation theory approaches. 
It is also noteworthy that the eigenfunctions of the TC Hamiltonian are invariant by orbital rotation, 
and therefore in the limit where one gives full flexibility to the wave function, 
the biorthogonal and usual orthogonal approaches give the same results. 
Nevertheless, as approximations are necessary made in the wave function ansatz in realistic systems, 
the biorthogonal approach might improve the quality of the results.

In latter works, Umezawa \textit{et. al.}\cite{UmeTsu-JCP-03,UmeTsuOhnShiChi-JCP-05} introduced a scheme to couple TC equations with the usual variational Monte Carlo (VMC) scheme in order to optimise the Jastrow factor through a variance miminization. 
An attempt to make TC equations variational have been proposed by Luo\cite{Luo-JCP-10,Luo-JCP-11} based on the empirical experience that in the original single determinant TC approach of Boys and Handy, the orbital optimization seemed to be the source of loss of the variational property. Luo therefore proposed to replace the original TC equations to optimize the orbitals by that of a general Jastrow-Slater ansatz.

Yanai and Chan

More recently, Alavi \textit{et. al}\cite{CohLuoGutDobTewAla-JCP-19} applied the TC equations to elaborate geminal functions explicitly taking into account for electron-electron correlation together with electron-electron-nucleus distance. The geminal function used in such work were obtained from the optimized Jastrow factors for He to Ne of Moskowitz \textit{et. al.}\cite{SchMos-JCP-90}. The main advantage of Alavi work relies on the fact that they allow a full flexibility to the Slater part of the wave function to adapt to the elaborated geminal functions. 
This is done through the use of a projective technique, the FCI-QMC, which allows to obtain the ground state of an operator, hermitian or not, through a stochastic sampling of the Hilbert space. Their approach share common points with that of Ten-No in the sense that they both use a fixed Jastrow factor and use a flexible form of wave functions, 
even though the FCIQMC is certainly more flexible than the coupled cluster ansatz of Ten-No used in Ref. \onlinecite{HinTanTen-JCP-01}.

The paper is organized as follows. 
In the first section, inspired by the short-range behaviour of RS-DFT effective wave functions,  
we derive the analytical form of a new Jastrow factor tuned by a unique parameter $\mu$, and discuss the physical content of such a simple Jastrow factor. 
We show then that the corresponding transcorrelated Hamiltonian has a straightforward analytical structure, which is briefly reviewed as a function of the $\mu$ parameter. 
Then, we perform a numerical study of the behaviour of the ground state eigenpair of $\tilde{H}[\mu]$ for the helium atom: 
in the first part we analyze the speed of convergence of the ground state energy with respect to the basis set and the parameter $\mu$, 
then we study the behaviour of the right eigenvectors in real space and compare it with the numerically exact ground state wave function. 
We show that the rapid convergence of the total energies coincides with a rapid convergence of the right eigenvectors in real space, and how to recover a very good approximation of the exact ground state wave function. 
Based on such encouraging results, we study the ground state energy of the helium isoelectronic series from Li$^+$ to Ne$^{8+}$ and compare different schemes to obtain 
a value of $\mu$ which depends on the system and still provides a fast convergence of the total energies. 
Eventually, we study the H$_2$ molecule and compare different regimes of correlation.

\section{A new form of Jastrow factor for transcorrelated Hamiltonian}
This section is dedicated to the derivation of a new form of Jastrow factor mimicking at short range the correlation effects of RS-DFT. 
For the sake of simplicity, we derive the transcorrelated Hamiltonian for the helium atom for a general Jastrow factor depending on the inter electronic distance.  
Then we focus on the behaviour at small $r_{12}$ of different flavours of Hamiltonians 
and the corresponding discontinuity in their eigenvectors. 
We browse three different types of Hamiltonians: the usual physical Hamiltonian, the transcorrelated Hamiltonian together with the effective Hamiltonian used in RS-DFT.
Based on such consideration, we propose a new Jastrow factor mimicking RS-DFT Hamiltonian at small $r_{12}$. 
Then, we derive the analytical form of the corresponding new transcorrelated Hamiltonians $\tilde{H}[\mu]$ in the case of a general $N$-electron system, and study its physical content as a  function of the parameter $\mu$.

\subsection{A simple physical picture: the case of the helium atom}
\subsubsection{The transcorrelated Hamiltonian for helium atom}
Let us write the Hamiltonian of the helium atom in the $ r_{12} = \big|{\bf r}_1 - {\bf r}_2 \big|$ coordinate 
\begin{equation}
 H  = h_c + \frac{1}{r_{12}},
\end{equation}
where 
\begin{equation}
 \begin{aligned}
 \label{def:h_c}
 h_c = &-\frac{1}{2} \sum_{i=1}^2 \bigg(\deriv{}{r_i}{2} + \frac{2}{r_i} \deriv{}{r_i}{} + \frac{2 Z}{r_i}\bigg) \\
     &-\bigg( \deriv{}{r_{12}}{2} + \frac{2}{r_{12}} \deriv{}{r_{12}}{} \bigg) \\
     &-\bigg( \frac{\bd{r_1}}{r_1} \cdot \frac{\bd{r_{12}} }{r_{12}}  \deriv{}{r_1}{} + 
                \frac{\bd{r_2}}{r_2} \cdot \frac{\bd{r_{21}} }{r_{21}}  \deriv{}{r_2}{} \bigg),
 \end{aligned}
\end{equation}
\label{sec:he_j}
Now let us consider the transcorrelated Hamiltonian $\tilde{H}[u]$ obtained by a similarity transformation of the usual Hamiltonian by a general Jastrow factor $u(r_{12})$ depending only on $r_{12}$: 
\begin{equation}
 \label{eq:ht_0}
 \begin{aligned}
 \tilde{H}[u]&= e^{-u(r_{12})} H e^{u(r_{12})}.
 \end{aligned}
\end{equation}
The only additional terms arising in $\tilde{H}[u]$ with respect to $H$ are those coming from the action of the differential operator in $r_{12}$,
\begin{equation}
 \mathcal{T}[u] =  -e^{-u(r_{12})}\bigg( \deriv{}{r_{12}}{2} + \frac{2}{r_{12}} \deriv{}{r_{12}}{} \bigg)e^{u(r_{12})}.  
\end{equation}
After some derivation, one can notice that by defining the following operators 
\begin{equation}
 \label{eq:def_tt}
 \tilde{t}[u] = -2 \deriv{u(r_{12})}{r_{12}}{} \deriv{}{r_{12}}{},
\end{equation}
\begin{equation}
 \label{eq:def_wt}
 \tilde{W}[u] = -\frac{2}{r_{12}} \deriv{u(r_{12})}{r_{12}}{}  , 
\end{equation}
\begin{equation}
 \label{eq:def_wt}
 \tilde{w}[u] = -\deriv{u(r_{12})}{r_{12}}{2} - \bigg( \deriv{u(r_{12})}{r_{12}}{} \bigg)^2, 
\end{equation}
one can write the Similarity transformed Hamiltonian  as
\begin{equation}
 \label{eq:ht_general}
 \tilde{H}[u] = h_c + \frac{1}{r_{12}}  + \tilde{W}[u] + \tilde{w}[u] + \tilde{t}[u] ,
\end{equation}
Therefore, one can see from Eq. \eqref{eq:ht_general} that the two electrons experience a modified scalar potential given by $\frac{1}{r_{12}}  + \tilde{W}[u] + \tilde{w}[u]$, and an additional differential operator $\tilde{t}[u]$, 
the latter making $\tilde{H}[u]$ non-hermitian.

\subsubsection{$r_{12} \rightarrow 0$ limits for different forms of Hamiltonians}
\label{sec:cusp}
Having established the analytical form of the transcorrelated Hamiltonian $\tilde{H}[u]$ for a general Jastrow factor $u(r_{12})$ in the case of the helium atom, one can then study the behaviour of $\tilde{H}[u]$ when $r_{12}\rightarrow 0$ and compare it to two other types of Hamiltonian: the usual physical Hamiltonian and that entering the RS-DFT framework. 
The analysis of the $r_{12}\rightarrow 0$ behaviour leads to the different condition that the eigenvectors of such operators must fulfill. For the sake of simplicity of the notations we focus on the ground state of these operators but the arguments are general for any bounded eigenvectors.

In the case of the usual Hamiltonian, the exact ground state wave function $\psiex$ must satisfy the Schroedinger equation in real space 
\begin{equation}
 H \psiex(\br{}_1,\br{}_2) = E_0 \psiex(\br{}_1,\br{}_2)\quad \forall (\br{}_1,\br{}_2).
\end{equation}
When looking at $r_{12}\approx 0$, all terms multiplying $\frac{1}{r_{12}}$ in $H \psiex(\br{}_1,\br{}_2)$ must remain finite, which, according to Eq.\eqref{def:h_c}, translates into
\begin{equation}
 \label{eq:cusp_0}
 \lim_{r_{12}\rightarrow 0}\bigg( \frac{2}{r_{12}} \deriv{}{r_{12}}{} -\frac{1}{r_{12}}\bigg) \psiex(\bd{r_1},\bd{r_2})  = c,\quad c<\infty 
\end{equation}
or equivalently by multiplying Eq. \eqref{eq:cusp_0} by $r_{12}$
\begin{equation}
 \lim_{r_{12}\rightarrow 0}\bigg( 2 \deriv{\psiex(\bd{r_1},\bd{r_2})}{r_{12}}{} -\psiex(\bd{r_1},\bd{r_2}) \bigg) = 0, 
\end{equation}
which translates into the famous cusp condition for antiparallel spins of Kato\cite{Kat-CPAM-57}, 
\begin{equation}
 \deriv{\psiex(\bd{r_1},\bd{r_2})}{r_{12}}{}\Bigr|_{r_{12}=0} = \frac{1}{2} \psiex(r_{12}=0). 
\end{equation} 
For an excellent pedagogical and general derivation of the cusp conditions, see Ref.\onlinecite{Tew-JCP-08}.

>>>>

\definecolor{darkgreen}{HTML}{009900}
\usepackage[normalem]{ulem}
\newcommand{\sphi}[1]{\hat{{\bf S}}_{#1}}
\newcommand{\overlap}[2]{\langle #1 | #2 \rangle}
\newcommand{\matelem}[3]{\langle #1 | #2 | #3 \rangle}
\newcommand{\deriv}[3]{\frac{\partial^{#3} #1}{\partial {#2}^{#3}}}
\newcommand{\bd}[1]{{\bf {#1}}}
\newcommand{\br}[0]{{\bf {r}}}
\newcommand{\bri}[1]{{\bf r}_{#1}}
\newcommand{\bs}[0]{{\bf {s}}}
\newcommand{\dr}[1]{\text{d}{\bf {#1}}}
\newcommand{\psiex}[0]{\Psi^{\text{ex}}_0}
\newcommand{\phiex}[0]{\Phi^{\text{ex}}_0}
\newcommand{\phimu}[0]{\Phi^{\text{ex},\mu}_0}
\newcommand{\phimub}[0]{\Phi^{\mathcal{B},\mu}_0}
\newcommand{\xhimub}[0]{X^{\mathcal{B},\mu}_0}
\newcommand{\psimub}[0]{\Psi^{\mathcal{B},\mu}_0}
\newcommand{\phiimub}[0]{\Phi^{\mathcal{B},\mu}_i}
\newcommand{\basis}[0]{\mathcal{B}}
\newcommand{\energyex}[0]{E^{\text{ex}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\identity}{\mathds{1}}
\newcommand{\mur}[1]{\mu({\bf r_{#1}})}
\newcommand{\muueg}{\mu_{\text{UEG}}}
\newcommand{\muuegav}{\langle \mu_{\text{UEG}}\rangle}
\newcommand{\mursc}{ \mu_{r_{s,c}}}
\newcommand{\murscav}{\langle \mu_{r_{s,c}}\rangle}
\newcommand{\mursclda}{\langle \mu_{r_{s,c}^{\text{UEG}}}\rangle}

@@

\maketitle
\section{Introduction}
Two formulation : the "pseudo orbital" (\textit{i.e.} usual common set of orbitals for bra and ket) and "biorthonormal orbitals" (\textit{i.e.} a different set of orbitals for bra and ket having the biorthonormal property). 
In the limit of the FCI wave function, these two formulations coincide, therefore we keep the "pseudo orbital" theory for the sake of simplicity.

\section{Brief overview of mathematical properties of similarity transformed Hamiltonian}
\subsection{Definition of the similarity transformation}
Assume an \textit{Hermitian operator} $H$ whose spectral representation reads 
\begin{equation}
 \label{eq:h_spectr}
 H  = \sum_n E_n \ket{\Psi_n}\bra{\Psi_n}
\end{equation}
with $\ket{\Psi_n}$ being its orthonormal eigenvectors 
\begin{equation}
 \bra{\Psi_n} \Psi_m \rangle = \delta_{mn},
\end{equation}
and $E_n \in \R$ its real eigenvalues.

Let us consider now an \textit{Hermitian} and \textit{invertible} operator $U$, \textit{i.e.} an operator $U$ such that there exists $U^{-1}$ such that 
\begin{equation}
 \label{eq:u_def_1}
 U^{-1} U = U \, U^{-1} = 1,
\end{equation}
and that 
\begin{equation}
 \begin{aligned}
 \label{eq:u_def_2}
 & U^\dagger = U \\
 & \big(U^{-1}\big)^{\dagger} = U^{-1}.
 \end{aligned}
\end{equation}
Then, the similarity transformed Hamiltonian $\tilde{H}$ is defined as
\begin{equation}
 \label{eq:def_htilde}
 \tilde{H} = U^{-1}\, H \, U.
\end{equation}
\subsection{Eigenvalues ans eigenvectors of $\tilde{H}$}
\subsubsection{Definition of left and right eigenvectors}
Regarding the properties of $\tilde{H}$, a first remark is that it is not Hermitian anymore
\begin{equation}
 \begin{aligned}
 \tilde{H}^\dagger & = \bigg(  U^{-1} \, H \,U \bigg)^\dagger \\
                   & = U\, H \, U^{-1}  \\
                   & \ne \tilde{H}.
 \end{aligned}
\end{equation}
Even though $\tilde{H}$ is not hermitian, it conserves the same eigenvalues than the original Hamiltonian $H$. 
The main difference with the original Hamiltonian are its eigenvectors: because $\tilde{H}$ is not Hermitian anymore, it possesses \textit{right} eigenvectors and \textit{left} eigenvectors. 
The right eigenvectors of a given operator $A$ are defined by
\begin{equation}
 A \ket{\Phi_n} = A_n \ket{\Phi_n}, 
\end{equation}
whereas the left eigenvectors of such operator are defined through
\begin{equation}
 A^{\dagger} \ket{X_n} = A_n \ket{X_n}.
\end{equation}
Therefore, if $A$ is hermitian, $A = A^{\dagger}$ and therefore $\ket{\Phi_n} = \ket{X_n} $, but for a non hermitian operator, the left and right eigenvectors do not coincide.

To find the right and left eigenvectors of $\tilde{H}$, one can insert the definition of $\tilde{H}$ (see Eq. \eqref{eq:def_htilde}) in Eq. \eqref{eq:h_spectr}
 \label{eq:tilde_1}
 \tilde{H} & = \sum_n E_n \, U^{-1} \, \ket{\Psi_n} \bra{\Psi_n}U.
We can define the right eigenvectors $\ket{\Phi_n}$ as
 \label{eq:def_right}
 \ket{\Phi_n} = U^{-1} \ket{\Psi_n},
 because  from Eq. \eqref{eq:tilde_1}, we notice that the $\ket{\Phi_n}$ fulfil the right eigenvector relationship
 \tilde{H} \ket{\Phi_n} & = \sum_m E_m \,U^{-1} \, \ket{\Psi_m} \bra{\Psi_m} U\ket{\Phi_n} \\
                        & = \sum_m E_m \,U^{-1} \, \ket{\Psi_m} \bra{\Psi_m} U \bigg( U^{-1} \ket{\Psi_n} \bigg) \\
                        & = \sum_m E_m \,U^{-1} \, \ket{\Psi_m} \bra{\Psi_m} \Psi_n \rangle \\
                        & = E_n          U^{-1} \ket{\Psi_n} \\
                        & = E_n  \ket{\Phi_n} \\
In a similar way, we can define the left eigenvectors $\ket{X_n}$ as 
 \label{eq:def_left}
 \ket{X_n} = U \ket{\Psi_n},
because they fulfil the left eigenvalue equation
 \tilde{H}^{\dagger} \ket{X_n}  = E_n \, \ket{X_n}. 
\subsubsection{Biorthogonality, spectral representation and loose of the variational principle}
The left and right eigenvectors fulfill the so-called \textit{biorthogonal} property, as 
from Eqs. \eqref{eq:def_right} and \eqref{eq:def_left} one obtains 
 \overlap{X_j}{\Phi_i} = \bigg(U \ket{\Psi_n} \bigg)^\dagger \bigg(U^{-1}\ket{\Psi_J}\bigg)
which, according to the Hermitian property of $U$ (see Eq.\eqref{eq:u_def_2}) simply becomes 
 \overlap{X_j}{\Phi_i} &= \matelem{\Psi_j}{U\,U^{-1}}{\Psi_i} \\
                       &= \overlap{\Psi_j}{\Psi_i} \\
                       &= \delta_{ji}.
Nevertheless, the right and left eigenvectors among themselves are not orthonormal as 
  \overlap{\Phi_j}{\Phi_i} & = \matelem{\Psi_j}{U^{-1}\,U^{-1}}{\Psi_i} \\
                           & = \matelem{\Psi_j}{U^{-2}}{\Psi_i},  
  \overlap{X_j}{X_i} & = \matelem{\Psi_j}{U\,U}{\Psi_i} \\
                           & = \matelem{\Psi_j}{U^{2}}{\Psi_i}.  
Having obtained biorthonormal left and right eigenvectors, one can then obtain a spectral representation of the similarity transformed Hamiltonian 
 \tilde{H} = \sum_i E_i \ket{\Phi_i} \bra{X_i}, 
which is very useful to study the variational principle, an important aspect of quantum chemistry. 
Let us consider a wave function $\Psi$ developed on the eigenvectors of $H$
 \ket{\Psi} = \sum_i c_i \ket{\Psi_i}.
The expectation value of $H$ over $\Psi$ is then 
  \matelem{\Psi}{H}{\Psi} & = \sum_{i,j} c_i c_j \matelem{\Psi_j}{H}{\Psi_i} \\
                          & = \sum_{i,j} c_i c_j  E_i \overlap{\Psi_j}{\Psi_i} \\
                          & = \sum_{i} |c_i|^2 E_i \\
                          & \ge E_0
which gives the variational principle because of the orthogonality of eigenvectors of an \textit{Hermitian} operator. 
Now let us decompose the same wave function $\Psi$ on the right-eigenvectors of $\tilde{H}$ 
 \ket{\Psi} = \sum_i \tilde{c}_i \ket{\Phi_i},
and consider the expectation value of $\tilde{H}$ over $\Psi$
  \matelem{\Psi}{\tilde{H}}{\Psi} & = \sum_{i,j} \tilde{c}_i \tilde{c}_j \matelem{\Phi_j}{\tilde{H}}{\Phi_i} \\
                                  & = \sum_{i,j} \tilde{c}_i \tilde{c}_j E_i \overlap{\Phi_j}{\Phi_i} 
which has no reason to be an upper bound to $E_0$ as $\overlap{\Phi_j}{\Phi_i} \ne \delta_{i,j}$. 
Therefore, we see that we cannot apply straightforwardly the variational principle 
to similarity transformed Hamiltonians, because of the lack of orthonormality of the right eigenvectors.

\section{A new form of Jastrow factor for transcorrelated Hamiltonian}
\subsection{The similarity transformed Hamiltonian}
Let us write the Hamiltonian of the He atom in the ${\bf r}_{12} = \big|{\bf r}_1 - {\bf r}_2 \big|$ coordinate 
\begin{equation}
 H  = h_c + \frac{1}{r_{12}},
\end{equation}
where 
\begin{equation}
 \begin{aligned}
 \label{def:h_c}
 h_c = &-\frac{1}{2} \sum_{i=1}^2 \bigg(\deriv{}{r_i}{2} + \frac{2}{r_i} \deriv{}{r_i}{} + \frac{2 Z}{r_i}\bigg) \\
     &-\bigg( \deriv{}{r_{12}}{2} + \frac{2}{r_{12}} \deriv{}{r_{12}}{} \bigg) \\
     &-\bigg( \frac{\bd{r_1}}{r_1} \cdot \frac{\bd{r_{12}} }{r_{12}}  \deriv{}{r_1}{} + 
                \frac{\bd{r_2}}{r_2} \cdot \frac{\bd{r_{21}} }{r_{21}}  \deriv{}{r_2}{} \bigg),
 \end{aligned}
\end{equation}
\label{sec:he_j}
Now let us consider the Similarity transformed Hamiltonian $\tilde{H}[j]$
\begin{equation}
 \label{eq:ht_0}
 \begin{aligned}
 \tilde{H}[j]&= e^{-j(r_{12})} H e^{j(r_{12})},
 \end{aligned}
\end{equation}
where $j(r_{12})$ is a general Jastrow factor depending only on $r_{12}$. 
Therefore, the only new terms arising in $\tilde{H}[j]$ with respect to $H$ are those coming from the differential operator in $r_{12}$,
\begin{equation}
 \mathcal{T}[j] =  -e^{-j(r_{12})}\bigg( \deriv{}{r_{12}}{2} + \frac{2}{r_{12}} \deriv{}{r_{12}}{} \bigg)e^{j(r_{12})}.  
\end{equation}
Let us compute the action of $\mathcal{T}[j]$ on a general function $\phi(\br{}_1,\br{}_2)$ 
\begin{equation}
 \begin{aligned}
 \mathcal{T}[j]\phi(\br{}_1,\br{}_2) & =- e^{-j(r_{12})} \bigg[ \deriv{}{r_{12}}{2} + \frac{2}{r_{12}} \deriv{}{r_{12}}{}\bigg]  e^{j(r_{12})} \phi(\br{}_1,\br{}_2) \\ 
                            &=  -\deriv{\phi(\br{}_1,\br{}_2)}{r_{12}}{2} - \frac{2}{r_{12}} \deriv{\phi(\br{}_1,\br{}_2)}{r_{12}}{} \\
  &- 2 \deriv{j(r_{12})}{r_{12}}{} \deriv{\phi(\br{}_1,\br{}_2)}{r_{12}}{} \\ 
  &- \bigg[ \frac{2}{r_{12}} \deriv{j(r_{12})}{r_{12}}{}  + \deriv{j(r_{12})}{r_{12}}{2} + \bigg( \deriv{j(r_{12})}{r_{12}}{} \bigg)^2\bigg] \phi(\br{}_1,\br{}_2). 
 \end{aligned}
\end{equation}
Now, by defining the following operators 
\begin{equation}
 \label{eq:def_tt}
 \tilde{t}[j] = -2 \deriv{j(r_{12})}{r_{12}}{} \deriv{}{r_{12}}{},
\end{equation}
\begin{equation}
 \label{eq:def_wt}
 \tilde{W}[j] = -\frac{2}{r_{12}} \deriv{j(r_{12})}{r_{12}}{}  , 
\end{equation}
\begin{equation}
 \label{eq:def_wt}
 \tilde{w}[j] = -\deriv{j(r_{12})}{r_{12}}{2} - \bigg( \deriv{j(r_{12})}{r_{12}}{} \bigg)^2. 
\end{equation}
one can write the Similarity transformed Hamiltonian  as
\begin{equation}
 \label{eq:ht_phi}
 \tilde{H}[j] = h_c + \frac{1}{r_{12}}  + \tilde{W}[j] + \tilde{t}[j] + \tilde{w}[j] ,
\end{equation}
ore more explicitly 
\begin{equation}
 \tilde{H}[j] = h_c + \frac{1}{r_{12}}\bigg( 1 - 2 \deriv{j(r_{12})}{r_{12}}{}\bigg) + \tilde{t}[j] + \tilde{w}[j] .
\end{equation}
$\tilde{H}[j]$, because of the first-order differential operator in $\tilde{t}[j]$, is then non-hermitian,   
and the effective potential $ \frac{1}{r_{12}}  + \tilde{W}[j] + \tilde{w}[j]$ in $\tilde{H}[j]$ only depends on the shape of $j(r_{12})$.

\subsection{$r_{12} \rightarrow 0$ limits for different forms of Hamiltonians}
\label{sec:cusp}
Having established the general form of the transcorrelated Hamiltonian $\tilde{H}[j]$, in the case of the helium atom, one can then study the behaviour of the latter when $r_{12}\rightarrow 0$ and compare it to two other types of Hamiltonian: the usual physical Hamiltonian and that entering the RS-DFT framework. 
The analysis of the $r_{12}\rightarrow 0$ behaviour leads to the different conditions that the eigenvectors of such operators must fulfill. For the sake of simplicity of the notations we focus on the ground state of these operators but the arguments are general for any bounded eigenvectors.

In the case of the usual Hamiltonian, the exact ground state wave function $\psiex$ must satisfy the Schroedinger equation in real space 
\begin{equation}
 H \psiex(\br{}_1,\br{}_2) = E_0 \psiex(\br{}_1,\br{}_2)\quad \forall (\br{}_1,\br{}_2).
\end{equation}
When looking at $r_{12}\approx 0$, all terms multiplying $\frac{1}{r_{12}}$ in $H \psiex(\br{}_1,\br{}_2)$ must remain finite, which, according to Eq.\eqref{def:h_c}, translates into
\begin{equation}
 \label{eq:cusp_0}
 \lim_{r_{12}\rightarrow 0}\bigg( \frac{2}{r_{12}} \deriv{}{r_{12}}{} -\frac{1}{r_{12}}\bigg) \psiex(\bd{r_1},\bd{r_2})  = c,\quad c<\infty 
\end{equation}
or equivalently by multiplying by $r_{12}$
\begin{equation}
 \lim_{r_{12}\rightarrow 0}\bigg( 2 \deriv{\psiex(\bd{r_1},\bd{r_2})}{r_{12}}{} -\psiex(\bd{r_1},\bd{r_2}) \bigg) = 0, 
\end{equation}
which translates into the famous cusp condition for antiparallel spins of Kato\cite{Kat-CPAM-57}, 
\begin{equation}
 \deriv{\psiex(\bd{r_1},\bd{r_2})}{r_{12}}{}\Bigr|_{r_{12}=0} = \frac{1}{2} \psiex(r_{12}=0). 
\end{equation} 
For a pedagogical and general introduction to the cusp conditions, see Ref.\onlinecite{Tew-JCP-08}.
